Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-04-25 joint_inf.tex
fiddling
update notation

With this split formulation, define mappings from patterns to splits as described above
$$
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet), \ \ancestralToSplit:\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet).
$$
Given a site-pattern-valued random variable $\alignmentColumnRV$ and an ancestral-state-valued random variable $\ancestralStateColumnRV$, define random variables
$$
\siteSplitRV := \patternToSplit(\alignmentColumnRV), \ \ancestralSplitRV := \ancestralToSplit(\ancestralStateColumnRV).
$$
We can now write the $i$th factor of \eqref{eq:full_likelihood} as
$$
P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t)=P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t)
$$
for some $j$ and $k$.
Taking the maximum across internal states yields
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &=
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
&= P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}
%EM Just noticed that eta doesn't have to be a real partition, because they don't have to be disjoint, right?
%das: I think they do have to be disjoint. The partition is into "most likely" and "not most likely," which is disjoint and exhausts the set. Am I missing something?
%EM Perhaps we're talking about partition in different senses. I was saying that eta_i can intersect with another eta_j. Let's chat IRL on Monday.
Given $(\tau, t)$ there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\begin{align*}
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi \mid \siteSplitRV=\siteSplit_j, \tau, t).
\end{align*}

\begin{align}
L_\nCols'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
%EM Indexing broken here: we are taking a product over i, but i doesn't appear in the expression.
    %das: This is intentional, though: it allows us to proceed to the following line, grouping factors by site pattern since their probabilities don't depend on $i$. I can try to think of how to make this more clear, but I thought it was a pretty key step to be able to pass to site patterns, and the reason for the machinery involved in defining ancestral state partitions.
%EM My suggestion is to use \psi(y_i) until you switch to a product indexed by the j's.
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
%EM As I fixed above and described in person, \ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) is not what you want here, because it will inflate the value by |\ancestralSplitPartition_j(\tau, t)| . This also appears below.
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ appears.

\paragraph{Example}
%EM My long term suggestion is for you to put this in example in the main text and the above in the appendix.
Consider a fixed, binary four-tip tree $\tau$.
The set of all possible character assignments can be described as
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips at which a character is assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
$$
\mathcal{P}(\siteSplitSet) = \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\}.
$$
Had we not identified character complements, we would need to consider the additional splits
$$
\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
$$
but the symmetry of the binary character model allows us to focus just on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
We cannot similarly identify complements in this case since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
%EM I dropped the \ancestralSplitPartition(\siteSplit, \tau, t) notation above. You can revert it if you like, but I prefer the bold eta notation as an ordered list of sets.
A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; define $\ancestralSplitPartition(\siteSplit, \tau, t)$ as the partition of $\mathcal{P}(\ancestralSplitSet)$ into those most likely ancestral splits for that particular site split, topology and set of branch lengths.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
The probabilities of observing ancestral splits for site split $\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
%EM Seems to me that this just requires one character change so would be p(1-p)^4.
%das: I think it's wrong, but in the other direction: two character changes to get to the differing internal state, then one between the two internal states.
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
%EM Have you told us in this example what the labeled topology is?
%das: Nope! I wanted to be simple, but it seems maybe too simple. Maybe I should move the Felsenstein/Farris bit to this example?
%Erick stopped here.
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
Here, the partition into most likely ancestral states will proceed as
$$
P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p) \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5
$$
where the partition is the single element $\ancestralSplitPartition_j(\tau, p)=\emptyset$.
For the site split $\{1,3\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^3(1-p)^2,
$$
where
$$
P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, p) \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3
$$
with the most likely partition $\ancestralSplitPartition_j(\tau, p)=\{\emptyset,\{1,2\}\}.$

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
Define \eqref{eq:site_pattern_likelihood} as
$$
L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau, t),\tau,t) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
$$
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau,t),\tau,t)
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \tau, t) \\
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log likelihood
$$
\ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV\in\ancestralSplitPartition_j(\tau, t) \mid \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} is written as
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau',t_1)},\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau^*,t_2)},\tau^*,\hat{t}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\widehat{\fullAncestralSplitPartitions(\tau',t_1)}\}$ and $\{\hat{t}_2,\widehat{\fullAncestralSplitPartitions(\tau^*,t_2)}\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.

@@

Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified} as
\begin{equation}
\label{eq:partial-likelihood}
\ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
We obtain an upper bound for the joint maximum of \eqref{eq:partial-likelihood}
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
    + \ell^p_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau,t)},\tau,\hat{t})
$$
using Gibbs's inequality
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\{\hat{t},\widehat{\fullAncestralSplitPartitions(\tau,t)}\} = \arg\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
$$
Similarly, we have a lower bound as
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
    + \ell^p_{\tau^*,t^*}(\widehat{\fullAncestralSplitPartitions(\tau,t)},\tau,\hat{t}).
$$

@@

    &\qquad + p_{2}  \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\
    &\qquad + p_{3}  \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\

@@

    &\qquad + (2-p_{1}-p_{2}-p_{12}-p_{23})\cdot\log(1+x') \\
    &\qquad + (p_{1}+p_{2}+p_{12}+p_{23})\cdot\log(1-x') \\
    &\qquad + (2-p_{3}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\
    &\qquad + (p_{3}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\

@@

    &\qquad + (3-p_{0}-p_{13}-p_{1}-p_{2}-p_{12}-p_{23})\cdot\log(1+x') \\
    &\qquad + (p_{2}+p_{3}+2(p_{1}+p_{3}+p_{12}+p_{23}))\cdot\log(1-x') \\
    &\qquad + (3-p_{0}-p_{13}-p_{3}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\
    &\qquad + (p_{2}+p_{3}+2(p_{12}+p_{23}+p_{123}))\cdot\log(1-y') \\

@@

a_{1} = 3-p_{0}-p_{13}-p_{1}-p_{2}-p_{12}-p_{23},

>>>>

With this split formulation, define mappings from patterns to splits as described above
\patternToSplit:\alphabet^\nSiteRows\rightarrow\mathcal{P}(\siteSplitSet), \ \ancestralToSplit:\alphabet^\nAncestralStateRows\rightarrow\mathcal{P}(\ancestralSplitSet).
Given a site-pattern-valued random variable $\alignmentColumnRV$ and an ancestral-state-valued random variable $\ancestralStateColumnRV$, define random variables
\siteSplitRV := \patternToSplit(\alignmentColumnRV), \ \ancestralSplitRV := \ancestralToSplit(\ancestralStateColumnRV).
We can now write the $i$th factor of \eqref{eq:full_likelihood} as
    P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) &= P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
    &= P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) \\
for some $j$ and $k$.
Taking the maximum across internal states yields
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\ancestralSplit_k \mid \tau, t) &=
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t) \\
&= P(\siteSplitRV=\siteSplit_j \mid  \tau, t)\cdot\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k  \mid \siteSplitRV=\siteSplit_j, \tau, t).
%EM Just noticed that eta doesn't have to be a real partition, because they don't have to be disjoint, right?
%das: I think they do have to be disjoint. The partition is into "most likely" and "not most likely," which is disjoint and exhausts the set. Am I missing something?
%EM Perhaps we're talking about partition in different senses. I was saying that eta_i can intersect with another eta_j. Let's chat IRL on Monday.
Given $(\tau, t)$ there exists an ordered list of sets $\fullAncestralSplitPartitions(\tau, t)=(\ancestralSplitPartition_1(\tau, t),\ldots,\ancestralSplitPartition_\nSiteSplits(\tau, t))$ such that any element $\xi_j$ of the $j$th component $\ancestralSplitPartition_j(\tau, t)$ satisfies
\max_{\ancestralSplit_k\in\mathcal{P}(\ancestralSplitSet)} \ P(\ancestralSplitRV=\ancestralSplit_k \mid \siteSplitRV=\siteSplit_j, \tau, t) &= P(\ancestralSplitRV = \xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t).
This allows us to write the likelihood in \eqref{eq:profile_likelihood} as a product over site patterns as opposed to sites, i.e.,
\begin{align}
L_\nCols'(\fullAlignment;\tau, t) &= \max_{\fullAncestralStates} \ L_\nCols(\fullAlignment;\fullAncestralStates, \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralStateColumn_i} \ P(\alignmentColumnRV=\alignmentColumn_i, \ancestralStateColumnRV=\ancestralStateColumn_i \mid \tau, t) \\
%EM Indexing broken here: we are taking a product over i, but i doesn't appear in the expression.
    %das: This is intentional, though: it allows us to proceed to the following line, grouping factors by site pattern since their probabilities don't depend on $i$. I can try to think of how to make this more clear, but I thought it was a pretty key step to be able to pass to site patterns, and the reason for the machinery involved in defining ancestral state partitions.
%EM My suggestion is to use \psi(y_i) until you switch to a product indexed by the j's.
                             &= \prod_{i=1}^{\nCols} \ \max_{\ancestralToSplit(\ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i), \ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \tau, t) \\
                             &= \prod_{i=1}^{\nCols} \ P(\siteSplitRV=\patternToSplit(\alignmentColumn_i) \mid \tau, t) \cdot \max_{\ancestralToSplit(\ancestralStateColumn_i)\in\mathcal{P}(\ancestralSplitSet)} P(\ancestralSplitRV=\ancestralToSplit(\ancestralStateColumn_i) \mid \siteSplitRV=\patternToSplit(\alignmentColumn_i), \tau, t) \\
                             &= \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t)\cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)} \label{eq:site_pattern_likelihood}
\end{align}
where $\nCols_j(\fullAlignment)$ is the number of columns in $\fullAlignment$ where the site split $\siteSplit_j$ appears.

\paragraph{Example}
%EM My long term suggestion is for you to put this in example in the main text and the above in the appendix.
\begin{figure}
    \centering
    \includegraphics[width=.45\textwidth]{unrooted_four_taxa}
    \caption{Four taxa tree}
\end{figure}

Consider the fixed, binary four-tip tree $\tau$ in Fig.~\ref{fig:four-taxa-tree}.
The set of all possible character assignments can be described as
\begin{align*}
\mathcal{P}(\{1,2,3,4\}) &= \{\emptyset, \{1,2,3,4\}, \{1\}, \{2,3,4\}, \{2\}, \{1,3,4\}, \{3\}, \{1,2,4\}, \\
                         &\qquad \{1,2\}, \{3,4\}, \{1,3\}, \{2,4\}, \{2,3\}, \{1,4\}, \{1,2,3\}, \{1,4\}\}.
\end{align*}
where each set indicates the tips at which a character is assigned the character $1$.
For example, $\emptyset$ is the labeling $0000$ and $\{1,3,4\}$ is the labeling $1011$.
Symmetry allows us to group adjacent pairs in $\mathcal{P}(\{1,2,3,4\})$ into equiprobable splits, letting $\siteSplitSet=\{1,2,3\}$.
The unique site splits, collapsing complements, are
$$
\mathcal{P}(\siteSplitSet) = \{\emptyset, \{1\}, \{2\}, \{3\}, \{1,2\}, \{1,3\}, \{2,3\}, \{1,2,3\}\}.
$$
Had we not identified character complements, we would need to consider the additional splits
$$
\mathcal{P}(\{1,2,3,4\}) \setminus \mathcal{P}(\siteSplitSet) = \{\{1,2,3,4\}, \{2,3,4\}, \{1,3,4\}, \{1,2,4\}, \{3,4\}, \{2,4\}, \{1,4\}, \{4\}\},
$$
but the symmetry of the binary character model allows us to focus just on the elements of $\mathcal{P}(\siteSplitSet)$.
This tree has two internal nodes with $\ancestralSplitSet=\{1,2\}$ and unique ancestral state splits
$$
\mathcal{P}(\ancestralSplitSet) = \{\emptyset, \{1\}, \{2\}, \{1,2\}\}.
$$
We cannot similarly identify complements in this case since, for $\siteSplit\in\mathcal{P}(\siteSplitSet)$,
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\siteSplit, \tau, t)\neq P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\siteSplit, \tau, t)
$$
in general.

For each site split $\siteSplit\in\mathcal{P}(\siteSplitSet)$, we maximize the likelihood over all $\ancestralSplit\in\mathcal{P}(\ancestralSplitSet)$.
%EM I dropped the \ancestralSplitPartition(\siteSplit, \tau, t) notation above. You can revert it if you like, but I prefer the bold eta notation as an ordered list of sets.
A maximum will occur at one of possibly several ancestral splits in $\mathcal{P}(\ancestralSplitSet)$; for the $j$th site split, define $\ancestralSplitPartition_j(\tau, t)\subset\mathcal{P}(\ancestralSplitSet)$ as the set of most likely ancestral splits for that particular site split, topology and set of branch lengths.
As a simple example, say all branch lengths correspond to a probability $p$ ($< 1/2$) of changing character along that branch.
The probabilities of observing ancestral splits for $\siteSplit_j=\emptyset$ are
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5,
$$
$$
%EM Seems to me that this just requires one character change so would be p(1-p)^4.
%das: I think it's wrong, but in the other direction: two character changes to get to the differing internal state, then one between the two internal states. Another set of eyes might make a liar out of me, though.
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\emptyset, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^3(1-p)^2,
$$
$$
%EM Have you told us in this example what the labeled topology is?
%Erick stopped here.
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\emptyset, \tau, t) =
p^4(1-p).
$$
The set of most likely ancestral states contains a single element, here $\ancestralSplitPartition_j(\tau, p)=\{\emptyset\}$.
Then, taking $\xi_j\in\ancestralSplitPartition_j(\tau, p)$ we have
$$
P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\emptyset, \tau, t) =
(1-p)^5.
$$
For $\siteSplit_j=\{1,3\}$ we have
$$
P(\ancestralSplitRV=\emptyset \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{1,2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3,
$$
$$
P(\ancestralSplitRV=\{1\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
P(\ancestralSplitRV=\{2\} \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^3(1-p)^2.
$$
In this case, the set of most likely ancestral states is $\ancestralSplitPartition_j(\tau, p)=\{\emptyset,\{1,2\}\}$, and, for $\xi_j\in\ancestralSplitPartition_j(\tau, p)$,
$$
P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\{1,3\}, \tau, t) =
p^2(1-p)^3.
$$

@@

We are interested in the properties of the objective function being maximized in \eqref{eq:site_pattern_likelihood} as data becomes plentiful.
Define \eqref{eq:site_pattern_likelihood} as
$$
L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau, t),\tau,t) = \prod_{j=1}^{\nSiteSplits} \ \left[P(\siteSplitRV=\siteSplit_j \mid \tau, t) \cdot P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j, \tau, t)\right] ^{\nCols_j(\fullAlignment)}.
$$
Assume our $\nCols$ observations were generated from a model with parameters $(\tau^*, t^*)$.
We have
\begin{align}
    \frac{1}{\nCols} \log L_\nCols''(\fullAlignment;\fullAncestralSplitPartitions(\tau,t),\tau,t)
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot  \log P(\siteSplitRV=\siteSplit_j, \ancestralSplitRV=\xi_j \mid \tau, t) \\
        &= \sum_{j=1}^\nSiteSplits \frac{\nCols_j(\fullAlignment)}{\nCols}\cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) +
            \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]
\end{align}
so that, in the limit of infinite data,
\begin{equation}
\frac{1}{\nCols} \log L_\nCols''(\fullAlignment;,\tau,t) \rightarrow \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*) \cdot [\log P(\siteSplitRV=\siteSplit_j \mid \tau, t) + \log P(\ancestralSplitRV=\xi_j \mid \siteSplitRV=\siteSplit_j , \tau, t)]. \label{eq:site_pattern_profile_likelihood_mean}
\end{equation}
Define the the divergence quantity
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\siteSplitRV=\siteSplit_j \mid \tau, t),
$$
and the partial log likelihood
$$
\ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \sum_{j=1}^\nSiteSplits P(\siteSplitRV=\siteSplit_j \mid \tau^*, t^*)\cdot\log P(\ancestralSplitRV=\xi_j \mid \siteSplit_j, \tau, t)
$$
so that \eqref{eq:site_pattern_profile_likelihood_mean} becomes
\begin{equation}
    \label{eq:log_likelihood_simplified}
    \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
Recall $(\tau, t)$ are estimands and $(\tau^*, t^*)$ are fixed, generating parameters.
For data generated from $(\tau^*, t^*)$, if
\begin{equation}
\label{eq:inconsistency_inequality}
\ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau',t_1),\tau',\hat{t}_1) > \ell_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau^*,t_2),\tau^*,\hat{t}_2)
\end{equation}
for $\tau'\neq\tau^*$ with $\{\hat{t}_1,\hat{\fullAncestralSplitPartitions}(\tau',t_1)\}$ and $\{\hat{t}_2,\hat{\fullAncestralSplitPartitions}(\tau^*,t_2)\}$ estimated by maximizing \eqref{eq:profile_likelihood}, we have an inconsistency.

@@

Our approach will be to seek bounds for the terms inside the maximization in \eqref{eq:log_likelihood_simplified} as
\begin{equation}
\label{eq:partial-likelihood}
\ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) = \shannonDivergence_{\tau^*,t^*}(\tau,t) + \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
\end{equation}
We obtain an upper bound for the joint maximum of \eqref{eq:partial-likelihood}
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \le
    \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
+ \ell^p_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau,t),\tau,\hat{t})
$$
using Gibbs's inequality
$$
\shannonDivergence_{\tau^*,t^*}(\tau,t) \le \shannonDivergence_{\tau^*,t^*}(\tau^*,t^*)
$$
and
$$
\{\hat{t},\hat{\fullAncestralSplitPartitions}(\tau,t)\} = \arg\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell^p_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t).
$$
Similarly, we have a lower bound as
$$
\max_{t,\fullAncestralSplitPartitions(\tau,t)} \ \ell_{\tau^*,t^*}(\fullAncestralSplitPartitions(\tau,t),\tau,t) \ge
    \shannonDivergence_{\tau^*,t^*}(\tau,\hat{t})
+ \ell^p_{\tau^*,t^*}(\hat{\fullAncestralSplitPartitions}(\tau,t),\tau,\hat{t}).
$$

@@

    &\qquad + p_{2}  \cdot\log((1+x')^2   (1+w')(1+y')(1-y')) \\
    &\qquad + p_{3}  \cdot\log((1+x')(1-x')(1+w')(1+y')^2) \\

@@

    &\qquad + (2-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x') \\
    &\qquad + (2-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\

@@

    &\qquad + (3-p_{0}-p_{13}-p_{1}-p_{3}-p_{12}-p_{23})\cdot\log(1+x') \\
    &\qquad + 2(p_{1}+p_{3}+p_{12}+p_{23})\cdot\log(1-x') \\
    &\qquad + (3-p_{0}-p_{13}-p_{2}-p_{12}-p_{23}-p_{123})\cdot\log(1+y') \\
    &\qquad + 2(p_{2}+p_{12}+p_{23}+p_{123})\cdot\log(1-y') \\

@@

a_{1} = 3-p_{0}-p_{13}-p_{1}-p_{3}-p_{12}-p_{23},
